# knitting SAS

As long as SAS is not in your `PATH` variable, you need to specify its full path
with the `engine.path` chunk option.

Now, if you prefer to use SAS for data analysis, this shows it is 
possible to write the analysis in SAS, and then run:

```{r build-example, eval=FALSE, echo=TRUE}
library(knitr)
knit("knittingSAS.Rmd")
```

`knitr` is powerful enough to allow for multiple input languages in one file.
Note that SAS was used in the first chunk, the following is the equivalent in R.

*Thank you* to Matt Mulvahill for running this example in Linux, (remove the engine.path from the chunk options if running under linux).

----------------------------------------------------

## Example: Regression

Here is an example of a simple regression in both SAS and R.  Yeah, that's right, *two* analysis languages, one markup language, *one* final output.  The data came from <a>http://www.ats.ucla.edu/stat/sas/modules/graph.htm</a>.

```{r plot_in_sas, engine='sas', engine.path = "C:\\Program Files\\SASHome\\SASFoundation\\9.3\\sas.exe", eval=TRUE, echo=TRUE, results='asis'}
DATA auto ;
  INPUT make $  mpg rep78 weight foreign ;
CARDS;
AMC     22 3 2930 0
AMC     17 3 3350 0
AMC     22 . 2640 0
Audi    17 5 2830 1
Audi    23 3 2070 1
BMW     25 4 2650 1
Buick   20 3 3250 0
Buick   15 4 4080 0
Buick   18 3 3670 0
Buick   26 . 2230 0
Buick   20 3 3280 0
Buick   16 3 3880 0
Buick   19 3 3400 0
Cad.    14 3 4330 0
Cad.    14 2 3900 0
Cad.    21 3 4290 0
Chev.   29 3 2110 0
Chev.   16 4 3690 0
Chev.   22 3 3180 0
Chev.   22 2 3220 0
Chev.   24 2 2750 0
Chev.   19 3 3430 0
Datsun  23 4 2370 1
Datsun  35 5 2020 1
Datsun  24 4 2280 1
Datsun  21 4 2750 1
;
RUN; 

proc reg data = auto;
model mpg = weight foreign;
run;

```


### in R

```{r, tidy = FALSE}
dat <- 
"AMC     22 3 2930 0
AMC     17 3 3350 0
AMC     22 . 2640 0
Audi    17 5 2830 1
Audi    23 3 2070 1
BMW     25 4 2650 1
Buick   20 3 3250 0
Buick   15 4 4080 0
Buick   18 3 3670 0
Buick   26 . 2230 0
Buick   20 3 3280 0
Buick   16 3 3880 0
Buick   19 3 3400 0
Cad.    14 3 4330 0
Cad.    14 2 3900 0
Cad.    21 3 4290 0
Chev.   29 3 2110 0
Chev.   16 4 3690 0
Chev.   22 3 3180 0
Chev.   22 2 3220 0
Chev.   24 2 2750 0
Chev.   19 3 3430 0
Datsun  23 4 2370 1
Datsun  35 5 2020 1
Datsun  24 4 2280 1
Datsun  21 4 2750 1"

auto <- read.table(col.names = c("make", "mpg", "rep78", "weight", "foreign"),
                   text      = dat)
auto$foreign <- factor(auto$foreign, 0:1, c("No", "Yes"))
str(auto)

fit <- lm(mpg ~ weight + foreign, data = auto)
summary(fit)

library(ggplot2)
ggplot(auto) + aes(x = weight, y = mpg, color = foreign) + 
  geom_point() + geom_smooth(method = "lm")

```

```{r}
print(sessionInfo(), local = FALSE)
```

